---
title: "Exploring the Palmer Penguins Dataset with ggplot2"
author: "Aanchal Saxena"
date: "2025-10-16"
output: html_document
---

## Introduction

This project explores the *Palmer Penguins* dataset, a well-known dataset often used as an alternative to the Iris dataset. 
It contains size measurements for three penguin species observed in the Palmer Archipelago, Antarctica.

The goal of this project is to demonstrate my ability to create and customize data visualizations using **ggplot2**, progressing from simple scatter plots to more detailed and informative visualizations.

## Plotting Evolution

```{r}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

For every chunk in this document — hide all warnings and messages.

### Loading Libraries 
```{r setup, message=FALSE, warning=FALSE}
# Install and load necessary packages
#install.packages("tidyverse")
library(tidyverse)
library(ggplot2)
library(palmerpenguins)

# View the dataset
head(penguins)
```

We’ll be using the **tidyverse** for data handling and **ggplot2** for visualization.  
The `penguins` dataset includes columns such as species, island, bill length, flipper length, body mass, and sex.

### Basic Plot  
```{r basic-plot}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g))
```

This basic scatter plot shows the relationship between **flipper length** and **body mass** for all penguins.

### Adding Trend Line
```{r smooth-plot}
ggplot(data = penguins) + 
  geom_smooth(mapping = aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g))
```

Adding a smooth trend line helps visualize the overall relationship between the two variables.

### Adding Color and Facets 
```{r color-facet-plot}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  facet_wrap(~species)
```

Color differentiates the three penguin species, and faceting allows for side-by-side comparison of each.

### Exploring Multiple Variables 
```{r facet-grid-plot}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  facet_grid(sex ~ species)
```

Using `facet_grid()`, we can compare male and female penguins across all species.

### Finished Plot 
```{r final-plot}
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  labs(title = "Palmer Penguins: Body Mass vs. Flipper Length",
       subtitle = "Sample of Three Penguin Species",
       caption = "Data collected by Dr. Kristen Gorman") +
  annotate("text", x = 220, y = 3500, label = "The Gentoos are the largest",
           color = "purple", fontface = "bold", size = 3, angle = 25)
```

This final version adds a **title**, **subtitle**, **caption**, and a **text annotation** highlighting an observation — showing mastery of ggplot customization.




